---
title: "EEDBI Exploratory Data - All Cohorts"
output:
  html_document:
    df_print: paged
---

Notes on cohorts, diagnosis distribution and missing data

## BGI
All BGI samples are "diagnosed" malnourished. All experimental prep data (librayPrepDate, sequencingDate, rnaIsolationDate) is missing. This is the only cohort with repeat measures (2 dissections per individual). biospySite is unknown (labeled as "Other"). Sample preparation - poly (A) capture and ribosome depletion.

## UVA
Samples sequenced from two sites - UVA, USA and Icddr,b, Bangladesh. Libraries prepared and sequenced in 3 batches (more or less in a site-specific manner, e.g. 2/3 batches UVA only, 1 batch BEED and UVA samples). Sample preparation - only ribosome depletion. 

## CCHMC 
Samples sequenced from two sites - CCHMC, USA and AKU (or SEEM), Pakistan. Libraries prepared and sequenced in 2 batches (not site-specific, e.g. CCHMC and SEEM samples prepared in both batches). Sample preparation - poly (A) capture and ribosome depletion. 

Miscoded samples removed. See [discussion](https://www.synapse.org/#!Synapse:syn22101481/discussion/threadId=7915&replyId=24576) for context.

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
library(synapser)
library(tidyverse)
library(readr)
library(sageseqr)
library(tibble)
synLogin(silent = TRUE)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```


```{r functions}
# This function is adapted from sageseqr::identify_outliers() to accept quantile normalized counts
fuse <- function(counts, clean_metadata, color, shape, z = 4){

PC <- stats::prcomp(counts,
  scale. = TRUE,
  center = TRUE)

# Plot first 2 PCs
data <- data.frame(SampleID = rownames(PC$rotation),
                   PC1 = PC$rotation[,1],
                   PC2 = PC$rotation[,2])

# Percentage from each PC
eigen <- PC$sdev^2
pc1 <- eigen[1]/sum(eigen)
pc2 <- eigen[2]/sum(eigen)

# Samples outside ellipse with radii defined as z SDs from the mean are
# outliers
outliers <- as.character(
  data$SampleID[
    c(
      which(
        ((data$PC1 - mean(data$PC1))^2)/((z*stats::sd(data$PC1))^2) +
          ((data$PC2 - mean(data$PC2))^2)/((z*stats::sd(data$PC2))^2) > 1
      )
    ),
    drop = TRUE
  ]
)

plotdata <- dplyr::left_join(
  data,
  tibble::rownames_to_column(clean_metadata, "SampleID")
) %>%
  dplyr::mutate(label = .data$SampleID) %>%
  dplyr::mutate(label = ifelse((.data$label %in% outliers), .data$label, NA))

p <- ggplot2::ggplot(plotdata, ggplot2::aes(x = .data$PC1, y = .data$PC2))

p <- p + ggplot2::geom_point(ggplot2::aes(
  color = .data[[color]],
  shape = .data[[shape]]
)
)

p <- p + ggforce::geom_ellipse(
  ggplot2::aes(
    x0 = mean(data$PC1),
    y0 = mean(data$PC2),
    a = z*stats::sd(data$PC1),
    b = z*stats::sd(data$PC2),
    angle = 0
  )
)

p <- p + sagethemes::scale_color_sage_d() +
  sagethemes::theme_sage() +
  ggplot2::theme(legend.position = "right") +
  ggplot2::geom_text(
    ggplot2::aes(label = .data$label),
    family = "Lato",
    size = 4,
    hjust = 0,
    na.rm = TRUE
  )

return(
  list(
    plot = p,
    outliers = outliers
  )
)
}
```


```{r define_files}
bgi_synid <- "syn23521999"
uva_synid <- "syn23538559"
cchmc_synid <- "syn23538349"
bgi_counts_synid <- "syn23536851"
uva_counts_synid <- "syn23538555"
cchmc_counts_synid <- "syn23353602"
composite_score_synid <- "syn25660423"
```

```{r get_data}
bgi <- read_csv(synGet(bgi_synid)$path) %>% 
  mutate(sequencingSite = "BGI") %>% 
  mutate(diagnosis = "Malnourished") %>% 
  mutate(siteName = "BGI") %>% 
  mutate(biopsySite = "Unknown")

bgi <- clean_covariates(
  bgi,
  factors = c("sampleId", "subjectId", "biopsyNumber", "testResult", "sex", 
              "sequencingSite", "diagnosis", "siteName", "biopsySite"),
  continuous = c("rin", "ageDays", "pct_pf_reads_aligned", "pct_coding_bases",
                 "pct_intergenic_bases", "pct_intronic_bases", "ribosomal_bases"),
  sample_identifier = "sampleId"
  )
  
uva <- read_csv(synGet(uva_synid)$path) %>% 
  mutate(sequencingSite = "UVA")

uva <- clean_covariates(
  uva, 
  factors = c("subjectId", "diagnosis", "sex", "siteName", "biopsySite",
              "libraryPrepDate", "sequencingDate"), 
  continuous = c("ageDays", "pct_pf_reads_aligned", "pct_coding_bases",
                 "pct_intergenic_bases", "pct_intronic_bases"), 
  sample_identifier = "subjectId"
)

cchmc <- read_csv(synGet(cchmc_synid)$path) %>% 
  mutate(sequencingSite = "CCHMC")

cchmc <- clean_covariates(
  cchmc,
  factors = c("subjectId", "diagnosis", "sex", "siteName", "biopsySite",
  "libraryPrepDate", "sequencingDate", "rnaIsolationDate"), 
  continuous = c("ageDays", "pct_pf_reads_aligned", "pct_coding_bases",
  "pct_intergenic_bases", "pct_intronic_bases", "pct_ribosomal_bases"),
  sample_identifier = "subjectId"
)
md <- dplyr::bind_rows(bgi, uva, cchmc)
# remove libraryPrepDate, sequencingDate, rnaIsolationDate, pct_ribosomal_bases (0 for all cohorts)
# biopsyNumber and testResult (specific to BGI), RIN and ribosomal_bases missing from UVA and CCHMC (?)
# subjectId NA for CCHMC and UVA as sampleId is unique
md <- select(md, -one_of("libraryPrepDate", "sequencingDate", "rnaIsolationDate", "pct_ribosomal_bases", "biopsyNumber", "rin", "testResult", "subjectId", "ribosomal_bases"))

#miscoded
rm <- c("B002-1A", "B002-2A", "BC11104", "BC21049")

md <- md[!(rownames(md) %in% rm),]

#load filtered counts from all dfs 
bgi_counts <- read_tsv(synGet(bgi_counts_synid)$path)
uva_counts <- read_tsv(synGet(uva_counts_synid)$path)
cchmc_counts <- read_tsv(synGet(cchmc_counts_synid)$path)

all_counts <- inner_join(bgi_counts, uva_counts)
all_counts <- inner_join(all_counts, cchmc_counts)
foo <- all_counts[,!(colnames(all_counts) %in% rm)]
```


|Sequencing Site |   Site  | No Intervention (Normal)| Nutritional Intervention + Malnourished|Other|
|:-----|-----:|---------------:|------------------------:|--------:|
|BGI|BGI   |               0|                        55|0|
|UVA|BEED  |               0|                       43|0|
|UVA|UVA   |               8|                        0|10|
|CCHMC|CCHMC |              23|                        0|2|
|CCHMC|SEEM  |               0|                       52|0|

```{r subset_samples_by_diagnosis}
md <- md %>% 
  filter(diagnosis %in% c("Malnourished", "Normal", "Nutritional Intervention", "Other"))
md$diagnosis <- droplevels(md$diagnosis)
```


```{r all_samples}
# check repeat measures of BGI (color by subjectId)
# split by 14 samples 
# one <- rownames(bgi[bgi$subjectId[1:14]])
# second <- rownames(bgi[bgi$subjectId[15:28]])
# 
# bgi_plot_one <- identify_outliers(bgi_counts[,one], bgi[one,], color = "subjectId", shape = "siteName", size = "rin")
# bgi_plot_two <- identify_outliers(bgi_counts[,second], bgi[second,], color = "subjectId", shape = "siteName", size = "rin")
# bgi_plot_one$plot
# bgi_plot_two$plot
```


```{r composite_scores}
composite_score <- read_csv(synGet(composite_score_synid)$path)

composite_score <- select(
  composite_score, 
  STUDYID, 
  SUBJIDO, 
  INTBRGL, 
  PANCELD, 
  INEPILY,
  GOBCELL,
  VILSARC,
  composite_cont_top5)

# create sample list to match composite socres to samples with transciptome data
samples <- rownames(md)
samples <- gsub("-1A|-2A", "", samples)
samples <- gsub("C-", "C", samples)
composite_score <- composite_score[composite_score$SUBJIDO %in% samples,]
# compute average 
avg_cs <- composite_score %>%
  filter(!is.na(composite_cont_top5)) %>% 
  select(SUBJIDO, composite_cont_top5) %>%
  group_by(SUBJIDO) %>% 
  summarize(mean_composite_top5 = round(mean(composite_cont_top5), 1))

#add mapping variable to md
md <- md %>% 
  mutate(mapping = rownames(md)) %>% 
  mutate(mapping = gsub("-1A|-2A", "", mapping)) %>% 
  mutate(mapping = gsub("C-0", "C0", mapping)) %>% 
  mutate(sampleId = rownames(md))

# join mean composite scores
md <- left_join(md, avg_cs, by = c("mapping" = "SUBJIDO")) %>% 
  select(-mapping) %>% 
  column_to_rownames(var = "sampleId")

md <- rownames_to_column(md, var = "sampleId")
md <- clean_covariates(md, continuous = c("pct_pf_reads_aligned","pct_coding_bases","pct_intergenic_bases","pct_intronic_bases"), factors = c("sex", "ageDays","sequencingSite","diagnosis","siteName","biopsySite"), sample_identifier = "sampleId")
```


```{r subset_counts_by_diagnosis}
all_counts <- all_counts[,c("feature", rownames(md))]
```

# PCA

## All samples
```{r all_sample_pca}
#remove missing counts
plot_counts <- column_to_rownames(all_counts, var = "feature")
filt_counts <- filter_genes(md, plot_counts, conditions_threshold = 0.5, conditions = "sex", cpm_threshold = 1)
normalized_counts <- limma::voom(filt_counts)
normalized_counts <- limma::normalizeBetweenArrays(normalized_counts$E, method = "quantile")
# plot <- identify_outliers(normalized_counts, md, color = "diagnosis", shape = "sequencingSite", size = "mean_composite_top5")
plot <- fuse(normalized_counts, md, color = "diagnosis", shape = "sequencingSite")
plot$plot
```
\n

Outliers, based on logCPM expression, are `r glue::glue_collapse(plot$outliers, ", ", last = " and ")`.

## CCHMC and UVA

```{r cchmc_uva}
sub_norm_counts <- rownames_to_column(as.data.frame(normalized_counts), var = "feature")
sub_norm_counts <- sub_norm_counts[,c("feature", rownames(md)[md$sequencingSite %in% c("CCHMC", "UVA")])]
plot_counts <- column_to_rownames(sub_norm_counts, var = "feature")

sub_md <- md[md$sequencingSite %in% c("CCHMC", "UVA"),]
plot <- fuse(plot_counts, sub_md, color = "diagnosis", shape = "sequencingSite")
plot$plot
```
\n

Outliers, based on logCPM expression, are `r glue::glue_collapse(plot$outliers, ", ", last = " and ")`.

